Distance Sampling in \proglang{R}

### 2016-01-25 paper.Rmd
Edited Intro Section 1

Distance sampling [@Buckland:2001vm; @Buckland:2004ts; @buckland2015distance] encompasses a suite of methods used to estimate the density and/or abundance of biological populations. Distance sampling can be thought of as an extension of plot sampling. Plot sampling involves selecting a number of plots (small areas) at random within the study area and counting the objects of interest that are contained within each plot. By selecting the plots at random we can assume that the density of objects in the plots is representative of the study area as a whole. One of the key assumptions of plot sampling is that all objects within each of the plots are counted. Distance sampling relaxes this assumption in that observers are no longer required to see and count everything.

In conventional distance sampling the observer is either stood at a point or moving along a line and we only assume they see everything that occurs exactly at that point or on the line. It is then expected that the further an object is from the sampler (the point or line) the less likely it is that the observer will see it. By recording the distance to each of the detected objects we can fit a model to describe the probability of detection given distance from the sampler, which we refer to as the ‘detection function’.  The detection function can be used to infer how many objects were missed and thereby produce estimates of density and/or abundance. (LM: perhaps move this? Further information on distance sampling, including its application, various extensions to deal with more complex situations and recent methodological developments can be found at http://distancesampling.org/dbib.html.)

The Windows program Distance (or "DISTANCE"; for clarity henceforth "Distance for Windows") was first released in November 1998 as Distance for Windows version 3.5 and was initially based on its predecessor “TRANSECT”. Since this time it has evolved to include various design and analysis features. Distance for Windows versions 5 onwards have included \proglang{R} packages as the analysis engines providing additional more complex analysis options than those offered by the original FROTRAN engines. (LM to finish: The current version of Distance links to 4 R packages... Development in R progressing rapidly so allowing users to access new tools faster?)

(LM Add part here to say what this paper describes… pathway to sustainability using R?) Further details of the history of Distance for Windows are included in @Thomas:2010cf.

(LM: still need to incorporate this)Here we show how to fit detection functions, perform model checking and selection, and estimate abundance in \proglang{R} [@rcore] using the package \pkg{Distance} [@distance-pkg]. \pkg{Distance} is a wrapper package around the more complex (and more powerful) \pkg{mrds} package [@mrds-pkg] and offers a subset of the analyses possible with that package.

(LT: Possibly edit to add the actual years.) The Windows program Distance (or "DISTANCE"; for clarity henceforth "Distance for Windows") has been available for many years to perform distance sampling analyses (and before that program "TRANSECT"). From version 5 of Distance for Windows, \proglang{R} packages have been included to perform some parts of an analysis [further details of the history of Distance for Windows are included in @Thomas:2010cf].

(LT: This won't mean much unless we've defined detection function first -- possibly need to do a little more work introducing the method and being explicit about where to go to find out more -- e.g., encyclopaedia articles, free online stuff, etc.  The concepts mostly are explained a little further down here, so possibly just some re-ordering needed.)

>>>>

Distance sampling [@Buckland:2001vm; @Buckland:2004ts; @buckland2015distance] encompasses a suite of methods used to estimate the density and/or abundance of biological populations. Distance sampling can be thought of as an extension of plot sampling. Plot sampling involves selecting a number of plots (small areas) at random within the study area and counting the objects of interest that are contained within each plot. By selecting the plots at random we can assume that the density of objects in the plots is representative of the study area as a whole. One of the key assumptions of plot sampling is that all objects within each of the plots are counted. Distance sampling relaxes this assumption in that observers are no longer required to see and count everything within selected plots.

Conventional distance sampling assumes the observer is either stood at a point or moving along a line and will only see everything that occurs exactly at that point or on the line. It is then expected that the further away an object is from the point or line (the sampler) the less likely it is that the observer will see it. By recording the distance to each of the detected objects we can fit a model to describe the probability of detection given distance from the sampler, which we refer to as the ‘detection function’.  The detection function can be used to infer how many objects were missed and thereby produce estimates of density and/or abundance.

The Windows program Distance (or "DISTANCE"; for clarity henceforth "Distance for Windows") can be used to fit detection functions to such distance sampling data. It was first released in November 1998 as Distance for Windows version 3.5 and was initially based on its predecessor “TRANSECT”. Since this time it has evolved to include various design and analysis features. Distance for Windows versions 5 onwards have included \proglang{R} [@rcore]  packages as the analysis engines providing additional more complex analysis options than those offered by the original FROTRAN engine. Further details of the history of Distance for Windows are included in @Thomas:2010cf.

As Distance for Windows becomes increasingly reliant on analyses performed in \proglang{R} and new methods are being developed at a rate where it is unfeasible to make all these available in the Windows interface, we are encouraging the use of our \proglang{R} packages directly. Until now those wishing to use \proglang{R} for straight forward distance sampling analyses would have had to negotiate the complex package \pkg{mrds} [@mrds-pkg] designed for mark-recapture distance sampling. \pkg{Distance} is a wrapper package around \pkg{mrds} making it easier to get started with basic distance sampling analyses in \proglang{R}. We demonstrate how to use \pkg{Distance} to fit detection functions, perform model checking and selection, and estimate abundance and/or density.
